Binning Strategies and Related Loss for Large Data

### 2015-04-30 Paper.Rnw
all but the last section

<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9). The binning origin at (0,0) leads to 9\\% less net spatial loss than (-9,-9); with losses of 391067 units and 420458 units, respectively.">>=

@@

For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers. \ktm{We define the \textit{origin offset} for each dimension, ($o_x$, $o_y$), such that 
%
\begin{eqnarray}\label{OriginOffset}
(\beta_{0,x},\beta_{0,y}) = (x_{(1)}, y_{(1)}) - (o_x, o_y),
\end{eqnarray}  
%
where $(\beta_{0,x},\beta_{0,y})$ is the bivariate bin origin, $x_{(1)}$ and $y_{(1)}$ are the minimum data value for each dimension, and $(o_x, o_y)$ is the origin offset. Thus the offset indicates the number of units in each dimension to shift the origin below the highest possible origin. }

\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}). }

@@

In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
%\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss. it's a bit confusing with the two minima floating around. It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
\ktm{Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm. Note that for simplicity, only origin offsets for with equal offset in each dimension are included in this simulation. Since the data resolution is two units in each dimension, we will pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees. The round glyphs indicate the origin offset where the net loss reached an absolute minimum. For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.

@@

The reduced binned data from spatial binning contains the center and count information for all bins. The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot. \ktm{Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups. If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading. This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. Our exploration of frequency loss will focus on frequency binning with at most ten bins. }

@@

\ktm{Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins. However, using both -- quantile log frequency binning -- is not a recommended approach. The frequency groups for quantile binning are invariant to the log transformation because a monotone one-to-one transformation will not alter groupings based on quantiles. In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.}

\ktm{Since the loss scales are not comparable, the choice is guided by desired interpretability. Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning. For standard log frequency binning, the shade is to be interpreted as an ordinal indicator of the groups based on equally spaced groupings of log bin frequencies, whereas for quantile frequency binning the shade denotes groups based on frequency quantiles. This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot in univariate visualization.}

>>>>

<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9). The binning origin at (0,0) leads to 9\\% less net spatial loss than (-9,-9); with losses of 391,067 units and 420,458 units, respectively.">>=

@@

For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers. We define the \textit{origin offset} for each dimension, as the tuple, ($o_x$, $o_y$), by which we shift the bivariate bin origin $(\beta_{0,x},\beta_{0,y})$ from the minimal values in $x$ and $y$:
%
\begin{eqnarray}\label{OriginOffset}
(\beta_{0,x},\beta_{0,y}) = (x_{(1)}, y_{(1)}) - (o_x, o_y),
\end{eqnarray}  
%
where $x_{(1)}$ and $y_{(1)}$ are the minimal data values in $x$ and $y$. %, and $(o_x, o_y)$ is the origin offset. 
Thus the offset indicates the number of units in each dimension to shift the origin below the highest possible origin.

It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units  minimizes the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the $X$ dimension and $\alpha_y$ units in the $Y$ dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).

@@

In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
%\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss. it's a bit confusing with the two minima floating around. It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm. Note that for simplicity, only origin offsets  with equal offsets in each dimension are included in this simulation. Since the data resolution  consists of one unit in each dimension, we  pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees. The round glyphs indicate the origin offset where the net loss reaches an absolute minimum in the simulation. For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.
%\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
For the bivariately skewed exponential data, the one unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.

@@

\hh{XXX please re-phrase the paragraph such that it reflects that we do not WANT to bin colors, but that this effectively happens when we look at a figure, because of the limitations of human perception. We use the binning to reflect what happens in practice. The binning then allows us to measure what kind of loss occurs because of the limitations of the human cognitive system.}

\ktm{The reduced binned data from spatial binning contains the center and count information for all bins. The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot. }
Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups. If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading. This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. Our exploration of frequency loss will focus on frequency binning with at most ten bins.

@@

Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins. However, using both -- quantile log frequency binning -- is not a recommended approach \hh{XXX not recommended is too strong. It actually does not help to do both, as can be seen in the figure}. The frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles. 
In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.

Since the loss scales are not comparable, the choice is guided by desired interpretation. 
Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning. For standard log frequency binning, the shade is to be interpreted as an ordinal indicator of the groups based on equally spaced groupings of log bin frequencies, whereas for quantile frequency binning the shade denotes groups based on frequency quantiles. This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot in univariate visualization.
